\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-05 NAR_DRAFT.tex
morning
finish methods

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $P(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have

\begin{equation}
  \label{eqn:1}
$$
  P(m,\nu) = p(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
$$
\end{equation}
where the density $\textrm{f}_{m,\nu}(x_i)$ can take several forms.

@@

\subsection{The mutect likelihood function}
\batcave has been designed as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or > 2.0)\cite{Cibulskis2013}.
\bkmcomment{need a little more time with this.}

\subsection{Tumor simulations.}
\bkmcomment{if $\pi$ works above, describe profile as $\pi$, and $\pi$ as linear combination of 3 signatures.}
We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
The simulated mutation rate was $\mu = 3\cdot10^{-6}$.
Variants were selected from TCGA and PCAWG variant files(dates).
Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were spiked to create tumors with Bamsurgeon with default parameters \cite{Ewing2015a},
and called with MuTect 1.1.7 \cite{Cibulskis2013} with the following parameters:

\subsection{Evaluation methods}
ROC
PRC
ICI. We used loess smoothing "loess" function in R with span = 0.5.
\begin{tiny}
\begin{verbatim}

@@

\subsection{Simulated data}
We generated six different tumor/normal pairs of two types, 3 each of 100X depth whole genomes and 500X whole exomes.
For each of the two types we simulated tumors with variants drawn from three different mutation profiles chosen to resemble melanoma (concentrated), a breast cancer (intermediate), and a lung cancer (diffuse)  as described in Materials \& Methods and Figure \ref{NAR-sigfig}.
These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.

@@

The estimated effective mutation rate, $\mu/\beta$, is in the range of 3e-7 for all simulated tumors.
We note that the effective mutation rate is lower than the simulated rate of 3e-6.
In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted. 
\bkmcomment{possible to quantify that this is very close to the actual insertion rate. Do it}.

@@

\begin{table*}[b]
\tableparts{%
\caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
\label{table:01}%
}{%
\begin{tabular*}{\textwidth}{@{}lllllllll@{}}
\toprule
Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
\\
& & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
\\
\colrule
100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
\\
100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
\\
100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
\\
500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .0585 & .0462
\\
500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .0587 & .0568 
\\
500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .0565 & .0576
\\
Griffith et al. 2015 & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
\\
Shi et al. 2018 & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
\\
\botrule
\end{tabular*}%
}
{$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
\end{table*}

>>>>

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
\begin{equation}
  \label{eqn:hypothesis}
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
\end{equation}
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $P(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have

\begin{equation}
  \label{eqn:1}
$$
  \mathrm{P}(m,\nu) = \mathrm{p}(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
$$
\end{equation}
where the density $\textrm{f}_{m,\nu}(x_i)$ can take several forms.

@@

\subsection{The likelihood function}
\batcave has been designed as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or 2.0) \cite{Cibulskis2013}.
MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results.
In addition, MuTect is particularly useful because it directly reports the log ratio of the of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
We have used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with

\begin{equation}
  \label{eqn:tlod}
    $$
    \mathrm{TLOD} = \frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}.
    $$
\end{equation}

The log posterior odds is the log likelihood ratio (\texttt{TLOD}) + the log prior odds, so we can compute the posterior odds in favor of the alternate hypothesis for a given substitution type as

\begin{equation}
  \label{eqn:computed_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{log}_{10}(\mathrm{logit}(\mathrm{p}(\mathrm{S_{i,j}}))},
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S_{i,j}})$ is the prior probability of a substitution of type $\mathrm{S_{i,j}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.
They suggest a threshold for the posterior odds of 2, and a prior per-base probability of mutation of $3\cdot10^{-6}$, leading to a \texttt{TLOD} threshold of 6.3.
We compute the prior odds for Mutect as
\begin{equation}
  \label{eqn:mutect_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6},
  $$
\end{equation}
since $\mathrm{log}_{10}(\mathrm{logit}(3\cdot10^{-6}) \approx -6$.

\subsection{Tumor simulations.}
We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
Tumors were simulated with three different mutation profiles, all of which include a contribution from COSMIC mutation signature 1 which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The "Concentrated" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
The "Intermediate" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is common in lung cancers.
The "Diffuse" profile is simulated with $\mathbf{\pi}$ set to an equal combination of the COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair.
The simulated mutation rate was $\mu = 3\cdot10^{-6}$.
Variants were selected from a combination of Cancer Genome Atlas (TCGA) and Pan-cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were spiked to create tumors with Bamsurgeon with default parameters \cite{Ewing2015a},
and called with MuTect 1.1.7 \cite{Cibulskis2013} with the following parameters:

\begin{tiny}
\begin{verbatim}

@@

\subsection{Simulated data}
We generated six different tumor/normal pairs of two types, 3 each of 100X depth whole genomes and 500X whole exomes.
For each of the two types we simulated tumors with variants drawn from three different mutation profiles chosen to resemble melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse)  as described in Materials \& Methods and Figure \ref{NAR-sigfig}.
These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.

@@

The estimated effective mutation rate, $\mu/\beta$, is in the range of 3e-7 for all simulated tumors.
We note that the effective mutation rate is lower than the simulated rate of 3e-6.
In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted.

@@

\begin{table*}[b]
\tableparts{%
\caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
\label{table:01}%
}{%
\begin{tabular*}{\textwidth}{@{}lllllllll@{}}
\toprule
Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
\\
& & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
\\
\colrule
100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
\\
100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
\\
100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
\\
500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .0585 & .0462
\\
500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .0587 & .0568 
\\
500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .0565 & .0576
\\
\citet{Williams2016} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
\\
\citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
\\
\botrule
\end{tabular*}%
}
{$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
\end{table*}
